---
output:
  html_document:
    keep_md: yes
---
# Percentage of natives across the US
Matthew Bender

July 29th, 2015

[GitHub](https://github.com/codebender)

[Twitter](https://twitter.com/5280code)

## Synopsis

In my [previous script](https://www.kaggle.com/codebender/2013-american-community-survey/colorado-transplants),
I determined how many Coloradans are "natives", meaning born in Colorado. 43% of
Coloradans are natives. My follow up question to that was, how does that compare
to all other states? So that is the goal of this script.  I plan to identify the
top and bottom "nativeness" states and see how Colorado compares.


### Basic Setup

#### Dependencies
```{r ECHO=FALSE}
library(readr)
library(plyr)
library(dplyr)
library(tidyr)
library(ggplot2)
library(maps)

options(scipen = 9)
```

#### Load Data
```{r ECHO=FALSE}
censusData <- read_csv('../input/pums/ss13pusa.csv')
birthPlaces <- select(censusData, ST, POBP, PWGTP)
rm(censusData)
gc()
censusDataB <- read_csv('../input/pums/ss13pusb.csv')
birthPlacesB <- select(censusDataB, ST, POBP, PWGTP)
birthPlaces <- rbind(birthPlaces, birthPlacesB)
rm(censusDataB)
rm(birthPlacesB)
gc()
dim(birthPlaces)
```

## Data Analysis

### Group by State & Birth Place
```{r}
ECHO=TRUE
birthPlaces <- group_by(birthPlaces, ST, POBP) %>%
  summarize(Total=sum(PWGTP)) %>%
  arrange(desc(Total))
```

### Calculate & Merge State Populations
```{r}
totalPopulations <- group_by(birthPlaces, ST) %>%
  summarize(totalPopulation=sum(Total))

birthPlaces <- join(birthPlaces, totalPopulations, by = "ST")
```

### Filter to only Natives of each state
```{r}
natives <- filter(birthPlaces, ST == POBP)
```

### Decode State codes
```{r}
stateCodeCSV = "POBP,region
001,Alabama
002,Alaska
004,Arizona
005,Arkansas
006,California
008,Colorado
009,Connecticut
010,Delaware
011,District of Columbia
012,Florida
013,Georgia
015,Hawaii
016,Idaho
017,Illinois
018,Indiana
019,Iowa
020,Kansas
021,Kentucky
022,Louisiana
023,Maine
024,Maryland
025,Massachusetts
026,Michigan
027,Minnesota
028,Mississippi
029,Missouri
030,Montana
031,Nebraska
032,Nevada
033,New Hampshire
034,New Jersey
035,New Mexico
036,New York
037,North Carolina
038,North Dakota
039,Ohio
040,Oklahoma
041,Oregon
042,Pennsylvania
044,Rhode Island
045,South Carolina
046,South Dakota
047,Tennessee
048,Texas
049,Utah
050,Vermont
051,Virginia
053,Washington
054,West Virginia
055,Wisconsin
056,Wyoming"

stateCodes <- read_csv(stateCodeCSV)

natives <- join(natives, stateCodes, by = "POBP") %>%
  select(region, population=totalPopulation, natives=Total)

natives$region <- tolower(natives$region)
```

### Calculate "Nativeness"
```{r}
natives$nativeness <- natives$natives / natives$population * 100

natives <- arrange(natives, desc(nativeness))
```


## Results

### States with Highest percentage of natives
```{r}
head(natives, n=10)
stateOrder <- reorder(natives$region, -natives$nativeness)
ggplot(natives[1:10,], aes(stateOrder[1:10], nativeness)) +
  geom_bar(stat="identity", fill='#0072B2', color='black') +
  scale_y_continuous(breaks = seq(0, 80, by = 10)) +
  theme(axis.text.x=element_text(angle=90))+
  labs(title = "Top 10 Native Population States",
       x = "State", y = "Native Percentage")
```

### States with the Lowest percentage of natives
```{r}
natives <- arrange(natives, nativeness)
head(natives, n=10)
stateOrder <- reorder(natives$region, natives$nativeness)
ggplot(natives[1:10,], aes(stateOrder[1:10], nativeness)) +
  geom_bar(stat="identity", fill='#0072B2', color='black') +
  scale_y_continuous(breaks = seq(0, 50, by = 10)) +
  theme(axis.text.x=element_text(angle=90))+
  labs(title = "Bottom 10 Native Population States",
       x = "State", y = "Native Percentage")
```

### Nativeness Map
```{r}
us_state_map = map_data('state')
us_state_map <- join(us_state_map, natives, by = 'region')

ggplot(us_state_map, aes(long, lat, group=group)) +
  geom_polygon(aes(fill = nativeness), colour="black") +
  scale_fill_gradientn(colours=rainbow(5)) +
  theme_light() +
  theme(strip.background = element_blank(),
        strip.text.x     = element_blank(),
        axis.text.x      = element_blank(),
        axis.text.y      = element_blank(),
        axis.ticks       = element_blank(),
        axis.line        = element_blank(),
        panel.border     = element_blank(),
        panel.grid       = element_blank(),
        legend.position  = "right") +
  xlab("") + ylab("") +
  labs(title = "How many people live in the state they were born?",
       fill = 'Percent of \nNatives')
```


## Conclusions

As I expected, Colorado is in the group of states with the lowest percentages of
natives, 8th lowest.  It was really surprising to see how low Nevada's native 
percentage is, 25.7%.  There were 8 states with a native percentage above 70%, 
Louisiana leading every state with 78%.
